# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r echo=TRUE, message=FALSE}

library(dplyr)
library(ggplot2)
library(scales)

activity_df <- read.csv('activity.csv')

# Transform date column from factor into Date format
activity_df <- mutate(activity_df,
                      date = as.Date(date)
                      )

sum(complete.cases(activity_df))
head(activity_df)
str(activity_df)
tail(activity_df)




```




## What is mean total number of steps taken per day?

For this part of the assignment, we ignore the missing values in the dataset.

The histogram below shows the total number of steps taken each day. 

```{r activityHistogram}


# Summarize steps taken by date
activity_by_day_df <- group_by(activity_df, date) %>%
  summarize(
    total_steps = sum(steps, na.rm = TRUE)
    )

# Plot histogram
ggplot(activity_by_day_df, aes(date, total_steps)) +
  geom_bar(stat = 'identity') +
  xlab('Day of month') +
  ylab('Number of steps') +
  ggtitle('Total number of steps taken each day by individual') +
  theme_bw() + 
  theme(plot.title = element_text(size = rel(1.2)),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 9),
        axis.text.x = element_text(angle = 90, hjust = 1,
                                   vjust = 1)) +
  scale_x_date(
    breaks = '1 day',
    limits = c(as.Date('2012-10-01'), as.Date('2012-11-30')),
    labels = date_format('%d-%b'),
    expand = c(0,0.6)
    )
  
```

   
```{r}

print(paste0('The mean total number of steps taken per day is: ',
             round(mean(activity_by_day_df$total_steps, na.rm = TRUE)),
             digits = 2))

paste0('The median total number of steps taken per day are: ',
       median(activity_by_day_df$total_steps, na.rm = TRUE))

```

The mean total number of steps taken per day are: `r mean(activity_by_day_df$total_steps, na.rm = TRUE)`

The median total number of steps taken per day are: `r median(activity_by_day_df$total_steps, na.rm = TRUE)`








## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
