# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r echo=TRUE, message=FALSE}

library(dplyr)
library(ggplot2)
library(scales)

activity_df <- read.csv('activity.csv')

# Transform date column from factor into Date format
activity_df <- mutate(activity_df,
                      date = as.Date(date)
                      )

```




## What is mean total number of steps taken per day?

For this part of the assignment, we ignore the missing values in the dataset.

The histogram below shows the total number of steps taken each day. 

```{r activityHistogram}


# Summarize steps taken by date
activity_by_day_df <- group_by(activity_df, date) %>%
  summarize(
    total_steps = sum(steps, na.rm = TRUE)
    )

#plot histogram
ggplot(activity_by_day_df, aes(total_steps)) + 
  geom_histogram(fill = 'steelblue', binwidth = 500) +
  xlab('Histogram of total steps by day') +
  ylab('Frequency') +
  ggtitle('Total number of steps taken each day') +
  theme_bw() +
  theme(plot.title = element_text(size = rel(1.2))
        )
```

   
```{r}

paste0('The mean total number of steps taken per day is: ',
             round(mean(activity_by_day_df$total_steps, na.rm = TRUE),
             digits = 2))

paste0('The median total number of steps taken per day are: ',
       median(activity_by_day_df$total_steps, na.rm = TRUE))

```


## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

Which 5-minute interval, on average across all the days in the dataset, contains the maximum 



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
